---
title: "Simple analysis for Twitter Data"
author: "Web Monitoring Group"
date: "11 October 2017"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analyse Twitter data  
We have a dataset of 500 tweets previously obtained using the *twitteR* package and by searching for the word "barcelona". We will first do look at the first few observations in the dataset.
```{r}
tweets <- readRDS("../data/barcelona-tweets.rds")
head(tweets)
```

We will routinely carry out some other operations just to have an overview or get a feel of what the dataset is like.

What are the dimensions of our data frame?
```{r}
(details <- dim(tweets))
```
There are `r details[1]` rows and `r details[2]` variables.

What are the names of our variables? We can list them so
```{r}
colnames(tweets)
```

Summary output of each variable
```{r}
summary(tweets)
```
This kind of summary can show some interesting results. For instance, looking at the variable `retweetCount`, wouldn't we love to know which tweet attracted 8,596 retweets? This can be easily done with R, but we won't go into that here...

Structure of the object i.e. the data frame
```{r}
str(tweets)
```

## Drawing a graph  
Although there are many native functions in R for drawing plots, We are going to be using a popular package known as *ggplot2* to draw our plots. We will learn about base R graphics another time.
```{r, message = FALSE}
library(ggplot2)
```

First, for the purpose of demonstration, we will load the dataset into a ggplot object, which will ultimately be used for plotting graphs. We will ask R to run this object (here called `gg`). When drawing a plot, we wouldn't normally run this step but it is being done here just to show what's going on.
```{r}
gg <- ggplot(tweets)
gg
```

All we see is a blank area, which means nothing. But we now have a bonafide **ggplot** object, which is the foundation of other features that we will be adding subsequently.

Let's now examine this object a bit for the sake of curiosity and better understanding
```{r}
typeof(gg)
```
```{r}
class(gg)
```
The `class` of an object is what makes object-oriented programming possible in R and developers can ascribe peculiar behaviours to an object. This is the secret behind the *ggplot2* package.

This object `gg` contains a whole lot of information including our dataset, layers, scales, mapping, theme, coordinates, etc.
```{r}
str(gg)
```

The elements of a list can be selected out using the `$` operator. For example, this object doesnot have any **layers** yet and when we check, it returns an empty list.
```{r}
str(gg$layers)
```

Now we're going to introduce the aesthetics function `aes()`, which is necessary to map variables from the dataset to parts of the plotting area.

To quickly demonstrate what this function does, check this out
```{r}
aes(created)
```
The above output shows that we are mapping our *created* variable to the x-axis of our plot, as can be seen below...
```{r}
gg <- ggplot(tweets, aes(created))
gg
```

We will now draw a simple graph of our tweets based on when they were created, precisely a density plot using `geom_density()`
```{r}
gg <- ggplot(tweets, aes(created)) + 
  geom_density()
gg
```

The `+` sign tells us that we are adding this on top of our plotting area as a layer. In the *ggplot2* package, you can add various elements of a plot one on top of another in this way, but we will not go into that here.

Now we are curious about what happened to the *layer* element of our `ggplot` object...
```{r}
str(gg$layers)
```

and it is now populated with a lot of information concerning our plot in its current state.

Now, we would like to explore the data in a multivariate fashion. We can check to see how different categories of tweets are distributed e.g. whether they are retweets or not (comparing `created` with `isRetweet` variable). We will do this by mapping another aesthetic, not to an axis, but to a colour scheme - by applying this to the function `geom_density()` that created the curve in the first place.
```{r}
gg <- ggplot(tweets, aes(created)) + 
  geom_density(aes(col = isRetweet))
gg
```

Let's try out the same thing with another categorical variable `truncated`. Hmm, what's in those extra long tweets?

We go back to disaggregating the tweets along the lines of whether or not they were retweets, and this time we want to shade the entire area under the curve by using a `fill` parameter instead.
```{r}
gg <- ggplot(tweets, aes(created)) +
  geom_density(aes(fill = isRetweet))
gg
```

Now let's introduce some transparency so that we can better visualise this using the **alpha** parameter.
```{r}
gg <- ggplot(tweets, aes(created)) + 
  geom_density(aes(fill = isRetweet), alpha = 0.5)
gg
```

We can draw a histogram of the number of times a particular item was retweeted and add a title to it.
```{r}
ggplot(tweets, aes(retweetCount)) +
  geom_histogram(bins = 100) +
  ggtitle(label = "Histogram of Retweet Counts")
```


Let us add another plot to our graph.  
```{r}
gg <- ggplot(tweets, aes(x = created, y = retweetCount)) +
  geom_point()
gg
```

Let us once again disaggregate this visualisation along the lines of whether they are **retweets*.
```{r}
gg <- ggplot(tweets, aes(x = created, y = retweetCount)) +
  geom_point(aes(col = isRetweet))
gg
```

Observe that each time we can simply add little pieces of code in order to make striking changes to the graphs.

To recap on ggplot:
1. Apply the dataset to `ggplot()`
2. Use at least one `aes()` function to determine the axes
3. Add other layers to 1. and 2. above (as your skill permits).

To be continued...